---
title: "IS606 Final Project"
author: "Rob Hodde"
date: "May 15, 2016"
output: html_document
---

```{r include=FALSE, cache=FALSE}
# DO NOT REMOVE
# THIS IS FOR SETTING SOME PLOTTING PARAMETERS SO THAT YOUR PLOTS DON'T TAKE UP TOO MUCH SPACE
# IF YOU WOULD LIKE TO CHANGE THESE, SEE HELP FILES ON THE par() FUNCTION
# OR ASK FOR HELP
library(knitr)
## set global chunk options
opts_chunk$set(fig.path='figure/manual-', cache.path='cache/manual-', fig.align='center', fig.show='hold', par=TRUE)
## tune details of base graphics (http://yihui.name/knitr/hooks)
knit_hooks$set(par=function(before, options, envir){
if (before && options$fig.show!='none') par(mar=c(4,4,.2,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
})
```

```{r echo=FALSE}
# load data

```


### Part 1 - Introduction:



Work is a means to earn the money required for food, clothing, shelter, healthcare, education, and all other goods and services we use.  It is also an avenue for developing social networks, enhancing individual responsibility, expressing creativity, and benefiting society as a whole.^^1





people need work - 
Meet people and expand their social networks.
Feel productive and valued.
Shape a personal identity.
Build a future.
Develop new skills

http://www.workwithoutlimits.org/individuals/why_work


health benefits.  unemployed are more likely to suffer from obesity and depression

http://theconversation.com/a-philosophers-view-the-benefits-and-dignity-of-work-829




Household net worth
Employment participation rate
Deficit
real wages
manufacturing jobs
all jobs
manufacturing output


immigrant population
imports
export



### Part 2 - Data:


```{r}
#import the most current polling information into a data table for analysis
require(XML)
myURL <- "http://www.census.gov/foreign-trade/balance/c5700.html#2012" 

tables <- readHTMLTable(myURL) # read the web page
n.rows <- unlist(lapply(tables, function(t) dim(t)[1]))  #format to table 
df <- tables[[which.max(n.rows)]]  #assign the data to a data frame

write.csv(df, file = paste(file_loc, "Trade.csv",sep = ""),row.names=TRUE, na="") 

```




### Part 3 - Exploratory data analysis:


### Part 4 - Inference:


### Part 5 - Conclusion: 


### References:


### Appendix (optional):

Remove this section if you don't have an appendix

1. http://benefitof.net/benefits-of-working/


